---
title: "HW2: Health Savings Experiments"
date: "06 September 2020"
output:
  pdf_document 
# Mac users may need to replace the previous line with the following
# uncommented lines:
#  pdf_document:
#    latex_engine: xelatex
---

To understand why those who are poor are constrained in their ability to save 
for investments in preventative health, two researchers designed a field 
experiment in rural Kenya in which they randomly varied access to four 
innovative saving technologies. By observing the impact of these various 
technologies on asset accumulation, and by examining which types of people 
who benefit most from them, the researchers were able to identify key barriers 
to saving.

They worked with 113 ROSCAs (Rotating Savings and Credit Associations). A ROSCA 
is a group of individuals who come together and make regular cyclical 
contributions to a fund (called the “pot”), which is then given as a lump sum 
to one member in each cycle. In their experiment, the researchers randomly 
assigned 113 ROSCAs to one of the five study arms. In this exercise, we will 
focus on three study arms (one control and two treatment arms). The data file, 
`rosca.csv` is extracted from their original data, excluding individuals who 
have received multiple treatments for the sake of simplicity.

Individuals in all study arms were encouraged to save for health and were asked 
to set a health goal for themselves at the beginning of the study.  In the first 
treatment group (*Safe Box*), respondents were given a box locked with a 
padlock, and the key to the padlock was provided to the participants. They were 
asked to record what health product they were saving for and its cost. This 
treatment is designed to estimate the effect of having a safe and designated s
torage technology for preventative health savings.  

In the second treatment group (*Locked Box*), respondents were given a locked 
box, but not the key to the padlock. The respondents were instructed to call the 
program officer once they had reached their saving goal, and the program officer 
would then meet the participant and open the *Locked Box* at the shop where the 
product is purchased. Compared to the safe box, the locked box offered stronger 
commitment through earmarking (the money saved could only be used for the 
prespecified purpose selected by the participant). 

Participants were interviewed again 6 months and 12 months later. In this HW, 
our outcome of interest is the amount (in Kenyan shillings) spent on preventative 
health products after 12 months.

Descriptions of the relevant variables in the data file `rosca.csv` are:

 Name                             Description
 -------------------------------------------------------------
 `bg_female`                      `1` if female, and `0` otherwise. This is a pre-treatment variable.
 `bg_married`                     `1` if married, and `0` otherwise. This is a pre-treatment variable.
 `bg_b1_age`                      Age at baseline. This is a pre-treatment variable.
 `encouragement`                  `1` if participant received encouragement only (control group), and    									                   `0` otherwise 
 `safe_box`                       `1` if participant received safe box treatment, and `0` otherwise
 `locked_box`                     `1` if participant received lock box treatment, and `0` otherwise
 `fol2_amtinvest`                 Amount invested in health products at time of the second follow up         
 `has_followup2`                  `1` if participant appears in second followup (after 12 months), and 									                       `0` otherwise
 

## Question 0
Run the following the code chunk below to load the data set and create a new variable `treatment` that takes the value `control` if receiving only encouragement, `safebox` if receiving a safe box,  and `lockbox` if receiving a locked box. We then designate that R should treat this new variable as a factor variable.

```{r}
library(tidyverse)
library(readr)
rosca <- read_csv("data/rosca.csv")

rosca <- rosca %>%
  mutate(treatment = case_when(
    encouragement == 1 ~ 'control',
    safe_box == 1 ~ 'safebox',
    locked_box == 1 ~ 'lockbox'
  ))

rosca$treatment <- as.factor(rosca$treatment)
```


\newpage
## Question 1 (5 points)

What is the specific causal question (or questions) the researchers aimed to answer (write separate questions for each treatment)? What are the potential outcomes (there are 3)? What are the hypothesized treatment effects?
For the Locked Box intervention group, what is their average missing counterfactual? How do your recommend  estimating it? How would you describe the internal validity of this study?
How many individuals are in the control group? How many individuals are in each of the treatment arms? Use a table to show the counts.


```{r}
table(rosca$treatment)


```
# Your Answer here:
What is the specific causal question (or questions) the researchers aimed to answer (write separate questions for each treatment)? 
  1. What is the impact of peoples health goal and savings, if they used a safebox relative to not using a safebox?
  2. What is the impact of peoples health goal and savings, if they used a lockbox relative to not using the lockbox?

What are the potential outcomes (there are 3)? 
  Control: No change in amount of money saved. 
  Lockbox: People save a lot of money using the LockBox
  Safebox: People save a lot of money using a SafeBox

What are the hypothesized treatment effects?
  The hypothesized treatment effect would be that people in the treatment group would save money and have good health. 

For the Locked Box intervention group, what is their average missing counterfactual? 
  Factual: Treatment Group
  MCF: Control Group 

How do your recommend  estimating it? 
  We would estimate it by looking at the control group. 

How would you describe the internal validity of this study?
  The internal validity is high because the experiment was very randomized. 

How many individuals are in the control group? How many individuals are in each of the treatment arms? Use a table to show the counts.
  control lockbox safebox 
      111     195     117 
\newpage
## Question 2 (6 points)

What are the drop-out rates (those for whom 12 month outcomes are missing) across the treatment and control conditions? Does the nature of either of the treatments (safebox or lockbox) suggest to you drop-out might be higher or lower in one group? Do the drop-out rates by treatment condition suggest that internal validity may be compromised? Why? Do they suggest external validity may be compromised? Why? *Hint: you can add, subtract, multiply, and divide tables in R.*
Subset the data (we suggest giving the subset data a new object name) so that it contains only participants who were interviewed at 12 months during the second followup.  We will use this subset for the subsequent analyses. How many participants are left in each group of this subset?  


```{r}

prop.table(table(rosca$has_followup2, rosca$treatment), margin = 2)
roscafu <- subset(rosca ,has_followup2 == "1")
table(roscafu$treatment)
```
# Your Answer here:
What are the drop-out rates (those for whom 12 month outcomes are missing) across the treatment and control conditions? 
       control    lockbox    safebox
  1 0.91891892 0.94358974 0.91452991
  
Does the nature of either of the treatments (safebox or lockbox) suggest to you drop-out might be higher or lower in one group? 
  The Lock Box has higher dropout rates. 

Do the drop-out rates by treatment condition suggest that internal validity may be compromised? Why?
  The validity is not affected because the rates are consistent. 

Do they suggest external validity may be compromised? Why?
  The internal validity is not compromised because the internal validity isnt. 

Subset the data (we suggest giving the subset data a new object name) so that it contains only participants who were interviewed at 12 months during the second followup.  
How many participants are left in each group of this subset?  

  control lockbox safebox 
      102     184     107 

\newpage
## Question 3 (5 points)

Does receiving a safe box or lockbox increase the average amount invested in health products relative to encouragement only? We focus on the outcome measured 12 months from baseline during the second follow-up `fol2_amtinvest`. First, describe the distribution of this outcome over all study participants. Then find the average amount invested (in Kenyan shilling) in health products across the treatment and control conditions. Then calculate the differences in the the mean of amounts invested in health products between each of the treatment arms and the control group. Briefly interpret the results. What are strengths and limitations of using averages (rather than a different summary statistic) to estimate impacts given the distribution of the outcome measure?

## Answer 3

```{r}

roscafu %>%
group_by(treatment) %>%
summarize(mean(fol2_amtinvest))

summary(rosca$fol2_amtinvest)
hist(rosca$fol2_amtinvest)
```

## Q3

Does receiving a safe box or lockbox increase the average amount invested in health products relative to encouragement only? 
  Yes it does, the mean values are higher than the control values. 

We focus on the outcome measured 12 months from baseline during the second follow-up `fol2_amtinvest`. First, describe the distribution of this outcome over all study participants. Then find the average amount invested (in Kenyan shilling) in health products across the treatment and control conditions. Then calculate the differences in the the mean of amounts invested in health products between each of the treatment arms and the control group. Briefly interpret the results. 

    Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
    0.0    40.0   100.0   322.2   490.0  5700.0      30 
    
    The distribution of the outcome is very distributed and it is more to the left. This data is a right skewed distribution. 

What are strengths and limitations of using averages (rather than a different summary statistic) to estimate impacts given the distribution of the outcome measure?

  The strength is that we can evaluate a large set of data. The limitations are that they hide disparities in our data. 

  

## Answer 3

\newpage
## Question 4 (8 points)

Examine the distribution of the pre-treatment variables - gender (`bg_female`), age (`bg_b1_age`) and marital status (`bg_married`). Are participants in the two treatment groups different from those in the control group with regard to each of these three variables?  What does the result of this analysis suggest about the internal validity of the findings you calculated in the previous question?  Also, if you think the internal validity has been compromised, state whether you hypothesize this would lead to over- or under-estimates of the treatment effects in this case. Please provide calculations and interpretations for each of the three pre-treatment variables separately.  


## Answer 4

```{r}
#gender
tapply(rosca$bg_female, rosca$treatment, mean)

```

```{r}
#age

tapply(rosca$bg_b1_age, rosca$treatment, mean)



```

```{r}
#marital status

tapply(rosca$bg_married, rosca$treatment, mean)



```

## Q4
Examine the distribution of the pre-treatment variables - gender (`bg_female`), age (`bg_b1_age`) and marital status (`bg_married`). Are participants in the two treatment groups different from those in the control group with regard to each of these three variables? 

  The participants in the two groups are not very distributed compared to each other. 

What does the result of this analysis suggest about the internal validity of the findings you calculated in the previous question?

  The internal validity is high because the groups didn't have a very distributed values. 

Also, if you think the internal validity has been compromised, state whether you hypothesize this would lead to over- or under-estimates of the treatment effects in this case. 

  I dont think it compromised because the distributions are not very far apart. 

Please provide calculations and interpretations for each of the three pre-treatment variables separately.

Gender: The gender has high percentages are they all are close to each other. 
Age: The age has very low percentages and they are close to each other. 
Marital: The marital has similar trends as to gender. 


##

\newpage
## Question 5 (6 points)

Does receiving a safe box or a locked box have different effects on the investment of *married* versus *unmarried women*? Compare the mean investment in health products among married women across the three treatment conditions. Then compare the mean investment in health products among unmarried women the three treatment conditions. Briefly interpret the results.  How does this analysis address any internal validity issues discussed in Question 4? 

## Answer 5

```{r}
roscafu %>%
  filter(bg_married == 1 & bg_female == 1) %>%
  group_by(treatment) %>%
  summarise(mean(fol2_amtinvest))

roscafu %>%
  filter(bg_married == 0 & bg_female == 1) %>%
  group_by(treatment) %>%
  summarise(mean(fol2_amtinvest))



```
## Q5

Does receiving a safe box or a locked box have different effects on the investment of *married* versus *unmarried women*? Compare the mean investment in health products among married women across the three treatment conditions. Then compare the mean investment in health products among unmarried women the three treatment conditions. Briefly interpret the results.  

  The two groups do have an affect on the mean of the different groups. The means for the married people are higher then the unmarried people. 

How does this analysis address any internal validity issues discussed in Question 4? 
  The internal validity is veru high because we have randomized the people in the groups. 
##5
